---
title: "SKC_fish_1"
output: html_document
date: "2023-10-24"
---

# DRAFT: Background of the vignette for fish

Members of the K’avi have fished the waters on their land for many years
to provide food for their community. To manage these waters, K'avi tribal fishery managers have been monitoring fish growth rates across the local water bodies in their community. 


However, many members of the community have begun to experience 
health problems associated with heavy metal poisoning. 
Historically, a factory used to be located upstream of the riverways fished by 
the K’avi, and waste from this factory was dumped into the nearby waters. 
Tribal fishery managers are now concerned that members of the community 
are being exposed to heavy metals through the fish that have been caught in these
waters. As a result, they now want to start recording the concentrations of methyl
mercury and selenium in the belly fat of fish across different sites in the area.
Their hope is to gain insight as to if the symptoms their community 
members are facing are connected to the fish being sourced from their waters. 


Once they collect these data, the tribal fishery managers will need to present 
their findings to members of the community so they can better understand the
public health crisis at hand. 


In this module, there are three main goals: 
1. Explore data related to fish growth and heavy metal concentrations.
2. Explore K'avi community health data.
3. Connect these datasets and produce a graph that could be used 
to communicate the fishery managers' findings to their community. 

--------------------------------------------------------------------------------
# WORKSHOP: Learning goals (More for us as a team, rather than for SKC students)

1. Students become more comfortable opening R and loading in data from an 
external data file. 
  - Considerations: do we want this file to be a csv? 
  
2. Students gain exposure to basic commands to explore their data before 
they begin to analyze it. 
  - Considerations: are these coded in baseR, do we want students to use tidyverse?
  
3. Students gain experience loading in packages (tidyverse and ggplot). 

4. Students gain exposure to working with different data types(numeric, characters)

5. Students gain exposure to basic summary statistic arguments in R to pull the 
mean, median, max, and min values for different variables in the dataset. 

6. Students become equipped with how to make a graph in R (tidyverse, ggplot) 
using variables pulled from a larger dataset. Can open a discussion point: 
how do we present data in a way that can help us understand the problem at hand?
(i.e. in this case, it may be plotting mercury concentrations found in fish 
along different sites, and showing it next to a graph of mercury found in people
across different sites)


--------------------------------------------------------------------------------
# PART 1: Loading in and exploring the data

Before we begin to analyze the data, we first need to upload our raw data into R. 
Once we do that, we can then explore...

**QUESTION: do we want students to jump in using tidyverse right away?

## First, we check and set our working directory(is this something we want students
to do?)

```{r set wd}
getwd()
list.files()
```


## Load in packages 

```{r load packages}

library(tidyverse)
library(dplyr)
library(ggplot2)
```



## Now we will load in the dataset for the physical fish data

```{r fish}
fish <- read.csv(.....)
```



## Exploring the structure data (again in dplyr/tidyverse or Base?). Look at the 
different columns, what they contain etc. 

```{r explore data structure}

```



## Explore data types (we will have at least numeric and character data)

```{r explore data types}

```


# PART 2: Pulling (subsetting) variables/columns of interest and running summary
stats on them that can inform us of the situation at hand. 


## Draft lead in text: According to the EPA, the maximum safe level of mercury to 
be consumed is around 0.1 mg/kg/day. Tribal fishery managers are interested in how
mercury levels in fish belly fat vary across different monitoring sites. They are 
also interested in if the amount of mercury in fish is connected to fish 
size. 

Below, we will analyze data points from the "fish" data set to try and answer 
these questions. 


## What is the mean amount of mercury found in fish across all sites? Across 
different sites? 

```{r mean amount of mercury}

```


## What is the median amount of mercury found in fish across all sites? Across 
different sites? 

```{r median amount of mercury}

```


## More to be added! But may be better to work on it as a whole group (vignette
and code). 

# PART 3: Working with ggplot to generate a graph of mercury concentrations in 
fish belly fat against some variable of interest (fish length, boxplots for 
fish age). 


## Introducing ggplot and the syntax? Give a commented out example maybe for 
Georgia to walk through (ggplot(data, aes(x,y) etc.)) ? 

```{r ggplot, Scatter Plot}

#Again, setting the data frame, this time we are creating a plot using the 
#fish_1998 data
# %>% is the way that we tell the computer what data we are drawing from
fish_1998 %>%
#ggplot is the function that we are using to create a plot. aes() is used to tell ggplot #what we want to see on each axis, in this case x= length and y= mercury.
  ggplot(aes(x = length, y = mercury)) +
#Since ggplot is just a function to create plots, we need to further specify
#what kind of plot it should create. We can do this by using the geom_point
#function. Because we are making a scatter plot, we use the function geom_point().
#We can also add a few more parameters to the graph, such as changing the color
#of the points (color = 'blue), or the size of them (size [2]).
  geom_point(color = 'blue', size = 2) +
#We can also add shapes to the plot. Here we added a dashed line to show where 
#the EPA's guideline for unsafe levels of mercury in fish would fall. To create 
#this we used the geom_hline function which simply creates a horizontal line at
#a y-intercept that you set. We can also adjust the color and appearance of the 
#line with linetype and color.
  geom_hline(yintercept = 4.5, linetype= "dashed", color = 'red')
```


```{r}
# For this plot, we want to see the distributions of lengths across the fish and how many #of them are above 60 cm. We'll start with a very similar order of steps as the scatterplot #we just made.
fish_1998 %>%
  ggplot(aes(y=length, x= "fish"))+
# x="fish" indicates that the x axis doesn't have a variable and is titled fish. This #allows us to look at the variation in lengths of fish independent of another variable.
  geom_hline(yintercept = 50, color= "red")+
  geom_jitter()
# the jitter plot is a variation on the scatterplot, but it separates the data so that we #can distinguish how many points are above 60 cm. To see the difference, try geom_point () #and then geom_jitter ()!
```





