---
title: "SKC_fish_1"
output: html_document
date: "2023-10-24"
---

# DRAFT: Background of the vignette for fish

Members of the K’avi have fished the waters on their land for many years
to provide food for their community. To manage these waters, K'avi tribal fishery managers have been monitoring fish growth rates across the local water bodies in their community. 


However, many members of the community have begun to experience 
health problems associated with heavy metal poisoning. 
Historically, a factory used to be located upstream of the riverways fished by 
the K’avi, and waste from this factory was dumped into the nearby waters. 
Tribal fishery managers are now concerned that members of the community 
are being exposed to heavy metals through the fish that have been caught in these
waters. As a result, they now want to start recording the concentrations of methyl
mercury and selenium in the belly fat of fish across different sites in the area.
Their hope is to gain insight as to if the symptoms their community 
members are facing are connected to the fish being sourced from their waters. 


Once they collect these data, the tribal fishery managers will need to present 
their findings to members of the community so they can better understand the
public health crisis at hand. 


In this module, there are three main goals: 
1. Explore data related to fish growth and heavy metal concentrations.
2. Explore K'avi community health data.
3. Connect these datasets and produce a graph that could be used 
to communicate the fishery managers' findings to their community. 

--------------------------------------------------------------------------------
# WORKSHOP: Learning goals (More for us as a team, rather than for SKC students)

1. Students become more comfortable opening R and loading in data from an 
external data file. 
  - Considerations: do we want this file to be a csv? 
  
2. Students gain exposure to basic commands to explore their data before 
they begin to analyze it. 
  - Considerations: are these coded in baseR, do we want students to use tidyverse?
  
3. Students gain experience loading in packages (tidyverse and ggplot). 

4. Students gain exposure to working with different data types(numeric, characters)

5. Students gain exposure to basic summary statistic arguments in R to pull the 
mean, median, max, and min values for different variables in the dataset. 

6. Students become equipped with how to make a graph in R (tidyverse, ggplot) 
using variables pulled from a larger dataset. Can open a discussion point: 
how do we present data in a way that can help us understand the problem at hand?
(i.e. in this case, it may be plotting mercury concentrations found in fish 
along different sites, and showing it next to a graph of mercury found in people
across different sites)


--------------------------------------------------------------------------------
# PART 1: Loading in and exploring the data

Before we begin to analyze the data, we first need to upload our raw data into R. 
Once we do that, we can then explore...

**QUESTION: do we want students to jump in using tidyverse right away?

## First, we check and set our working directory(is this something we want students
to do?)

```{r set wd}
getwd()
setwd()
list.files()
```


## Load in packages 

Think of packages as tools for you to use in R that aren't automatically part of
R when you first installed it. The packages below allow you to do way more things
in R, and for coding to be much easier! These packages actually work in harmony 
with one another, and are part of the "tidyverse".

```{r load packages}
library(tidyverse)
library(dplyr)
library(ggplot2)
```
#Troubleshoot: If that didn't work, it might be because you don't already have 
#these packages installed. To install them, copy this code: install.packages("tidyverse") 
#above the line library(tidyverse), and see if that works. 


## Now we will load in the dataset for the physical fish data

```{r fish}
fish <- read.csv("fish_1998.csv", header = TRUE, sep = ',')
```
#skip this part for now, this is all for if/when we change the .RData file into a csv!
Let's break down the different parts of what just happened:
1) We told R to read the data (a .csv file) using the command read.csv()
2) Within the (""), we identify the data file from our working directory that we want to load into R
3) header = TRUE just tells R that the first row of data is the column names. The alternative is header = FALSE
4) sep is short for the word "separate", and it tells R that each column in the csv is separated by a comma ','
3) "fish" is the name we are giving that file in R, and is how we will refer to this data going forward.
4) <- is like = in a math equation, it tells R that "fish" is the same as the data "fish_1998.csv"

## Viewing the data
```{r view data pt. 1}
View(fish) 
```

## Different way to view the data
``` {r view data pt. 2}
head(fish) #just shows the column names and the first few rows of the data
```

What data do we have? What sort of questions can we ask with this data?

## Exploring the dimensions of the data

```{r explore data dimensions}
dim(fish) #shows the dimensions of the dataset, which will be important later on
```


# PART 2: Running some basic statistics! 

According to the EPA, the maximum safe level of mercury to be consumed is around
0.1 mg/kg/day. Tribal fishery managers are interested in how mercury levels in 
fish belly fat vary across different monitoring sites. They are 
also interested in if the amount of mercury in fish is connected to fish size. 

Below, we will analyze data points from the "fish" data set to try and answer 
these questions. 


## What is the mean amount of mercury found in fish across all sites? Across 
different sites? 

```{r mean amount of mercury}
mean(fish$mercury_1998)
#you can see that $ is how you tell R what variable/column you want to work with
```


## What is the median amount of mercury found in fish?

```{r median amount of mercury}
median(fish$mercury_1998)
```

#How would we interpret these median and mean values? Should the K'avi community be concerned?



## Now figure out what the maximum and minimum amount of mercury was found in fish:

``` {r max and minimum amount of mercury}


```

#How would we interpret these maximum and minimum values? Are they as important as the average?

#There is a faster way to get all of these basic statistics at once:
``` {r summary stats}
summary(fish$mercury_1998)
```

# PART 3: Working with ggplot to generate a graph of mercury concentrations in 
fish belly fat against some variable of interest (fish length, boxplots for 
fish age). 


## Introducing ggplot and the syntax? Give a commented out example maybe for 
Georgia to walk through (ggplot(data, aes(x,y) etc.)) ? 

```{r ggplot}







```






