---
title: "SKC_fish_1"
output: html_document
date: "2023-10-24"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DRAFT: Background of the vignette for fish

Members of the K’avi have fished the waters on their land for many years
to provide food for their community. To manage these waters, K'avi tribal fishery managers have been monitoring fish growth rates across the local water bodies in their community. 

However, many members of the community have begun to experience 
health problems associated with heavy metal poisoning. 
Historically, a factory used to be located upstream of the riverways fished by 
the K’avi, and waste from this factory was dumped into the nearby waters. 
Tribal fishery managers are now concerned that members of the community 
are being exposed to heavy metals through the fish that have been caught in these
waters. As a result, they now want to start recording the concentrations of methyl
mercury in the belly fat of fish in the area.The tribe uses guidance from the Environmental Protection Administration (EPA), who have determined that concentrations of methylmercury greater than 4 mg of mercury per kg of fish (**4 mg/kg**) pose health hazards. 

**NOTE TO SKC/ND DEVELOPERS: WE NEED TO CHECK THIS NUMBER.**

Back in 1998, the tribal fisheries managers caught fish and measured their length and also the concentration of mercury in the fish. The goals of these managers was to learn how common it was for concentrations of mercury to be above the safety level of 4 mg/kg.

Once they collect these data, the tribal fishery managers will need to present 
their findings to members of the community so they can better understand the
public health crisis at hand. 

In this module, there are three main goals: 

1. Explore data related to fish growth and heavy metal concentrations.
2. Explore K'avi community health data.
3. Connect these datasets and produce a graph that could be used 
to communicate the fishery managers' findings to their community. 

**NOTES TO SKC/ND DEVELOPERS:** 

Learning goals for this module:

1. Students become more comfortable opening R and loading in data from an 
external data file. 
  
2. Students gain exposure to basic commands to explore their data before 
they begin to analyze it. 

3. Students gain experience loading in packages (tidyverse and ggplot). 

4. Students gain exposure to working with different data types(numeric, characters)

5. Students gain exposure to basic summary statistic arguments in R to pull the 
mean, median, max, and min values for different variables in the dataset. 

6. Students become equipped with how to make a graph in R (tidyverse, ggplot) 
using variables pulled from a larger dataset. Can open a discussion point: 
how do we present data in a way that can help us understand the problem at hand?
(i.e. in this case, it may be plotting mercury concentrations found in fish 
along different sites, and showing it next to a graph of mercury found in people
across different sites)

**END NOTE TO DEVELOPERS**

# PART 1: Loading in and exploring the data

Before we begin to analyze the data, we first need to upload our raw data into R. 
Once we do that, we can then explore...

**Clearing Environment**

The first thing we do anytime we start working in R is cleaning our environment. Your environment is where R stores all of the objects it is working with since the last time you cleared your environment. We clear the environment each time we open R so that there isn't anything still open from the list time you were using R that will mess up what we are trying to do this session. You can clear your environment with the following code or by clicking the broom icon that can normally be found on the top right section of RStudio. 

**NOTE TO SKC/ND DEVELOPERS:** 
**THIS AND THE SECTION ON LOADING PACKAGES WILL ALSO BE COVERED IN MODULE 0**

```{r}
rm(list = ls())
```


## Setting the Working Directory

In the following code we will tell the computer where we want it to look for 
  files on our computer. This is called a working directory. We will discuss this
  more in depth later. --- Hannah: or will we? Did we do this in module 0? ---
```{r set wd}
current_path = rstudioapi::getActiveDocumentContext()$path
setwd(dirname(current_path))
```



## Load in packages 

"Packages" are computer code that someone wrote to do a particular thing. By loading a package, R downloads that code, so that you can do that thing yourself. Think of packages as tools for you to use in R that aren't automatically part of R when you first installed it. The package below allows you to do  more things in R, and makes coding much easier! 

Below, we install the package "tidyverse" which we'll use to organize data and to make pretty graphs.
 
```{r load packages}
# This line of code helps you install a package. 
# You only need to do this once on your computer
if(!require("tidyverse")) install.packages("tidyverse") 

# This loads a package that has already been installed, so you can use it.
library(tidyverse)
```


**Now we will load in the dataset for the physical fish data**

These data are the length and methyl mercury content of fish caught on the K'Avi reservation in 1998, soon after the polluting factory was shut down. 

* Each row is data from one fish. 
* The first column "length" is the length of the fish in cm. 
* The second column "mercury" is the concentration of methylmercury (a poison) in the fish, in units of milligrams of methylmercury per kilogram of fish (mg/kg). 

```{r fish_1998}
fish_1998 <- read.csv("fish_1998.csv", header = TRUE, sep = ',')
```

Let's break down the different parts of what just happened:
1) We told R to read the data (a .csv file) using the command read.csv()
2) Within the (""), we identify the data file from our working directory that we want to load into R
3) header = TRUE just tells R that the first row of data is the column names. The alternative is header = FALSE
4) sep is short for the word "separate", and it tells R that each column in the csv is separated by a comma ','
3) "fish" is the name we are giving that file in R, and is how we will refer to this data going forward.
4) <- is like = in a math equation, it tells R that "fish" is the same as the data "fish_1998.csv"
5) This command turns a file on your computer (fish_1998.csv) into a dataframe (fish_1998) that you can examine in RStudio. A 'dataframe' is the way R can see data that has rows and columns. When you run line of code above, R looks for the .csv file called "fish_1998.csv" and imports it into this RStudio session in a format (a database) that you can use for coding. The new dataframe (fish_1998) should appear in the RStudio "Environment" window in the upper right.

**NOTE to ND/SKC developers: Below is an alternative format for loading data in RData format (commented out for now)**

```{r fish_1998 RData}
# load("fish_1998.RData")
# fish_1998 <- fish_1998
```

**Viewing the data**

**Note:** This will open a spreadsheet in new tab in this window (the "source editor" window). 

* Look through these data to see if they make sense (any weird numbers? Any missing data?).
* Then close that tab and come back to this window.
```{r view data pt. 1}
View(fish_1998) 
```

**A shorthand way to view the data**

If we don't want to look at all of the data and instead just want to see what is going on in the first couple rows we can use the following code. 

The command head() shows the column names and the first few rows of the data

``` {r view data pt. 2}
head(fish_1998) 
```
**Class Questions:**

* What are the columns in the R dataframe called fish_1998?
* What sorts of questions could the tribal fishery managers answer with this data?

**Exploring the dimensions of the data**

To answer these questions, it would be helpful to know how many fish have methyl mercury measurements.We can very quickly find that out with the following code.

The command dim() shows the dimension of the dataset. "Dimension" is the number of rows & number of columns.
```{r explore data dimensions}
dim(fish) #shows the dimensions of the dataset, which will be important later on
```

**Class Question:**

* How many fish did the managers measure methylmercury on in 1998?

**Recap**
You now know the basic ways to look at and understand your data in R, which will
come in handy any time you use R in the future.

We now know what sort of questions we can ask with this data. Next we'll learn how can we use R to answer those questions?

# PART 2: Calculating statistics that describe the levels of mercury in fish in 1998 

According to the EPA, the maximum safe level of mercury to be consumed is around
4 mg/kg/day. 

Tribal fishery managers are interested in whether the mercury levels in 
local fish belly fat exceeds this amount. They are also interested in if the 
amount of mercury in fish is connected to fish size. 

Below, we will analyze data points from the "fish" data set to try and answer 
these questions. 


**What is the mean (average) amount of mercury found in local fish? **

```{r mean amount of mercury}
# Below, we will use "$" to select the column called "mercury" in the data
# If we wanted the mean fish length, we;d have written mean(fish_1998$length)

mean(fish$mercury)

```
The "mean" is the average level of mercury in the data.

**Class Questions:**

* Why would knowing the mean (average) amount of methylmercury in fish be useful for fishery managers and the community?

**The median**

The "median" is the middle level of mercury data (half the fish in the dataset have more mercury and half have less). 

* Means and medians are often similar, However:
  * They will be different if the data have a lot of large values or a lot of small valued (skewed data).
  * Sometimes you are asked to use one or the other, so it's good to be careful about which you are calculating.

```{r median amount of mercury}
# remember, in the last code chunk, we used mean(fish_1998$mercury)
# The code below is similar, but it calculates the median.

median(fish_1998$mercury) 
```
**Class Question:**

* How would we interpret these median and mean values? Should the K'avi community be concerned?


**Now figure out what the maximum and minimum amount of mercury was found in fish:**

**Class Question:**

Work in groups to see if you can find these values using the commands max() and min() 

For help, you can google "R min()" or you can type ?min() in the "Console" window below, then read the help file in the "Output" window on the lower right.

``` {r max and minimum amount of mercury}
# Type your code below to calculate the mininum and maximum values of methylemercury in the 1998 fish dataset.


```

**Class Question**

How would we interpret these maximum and minimum values? Are they as important as the average?

**A fast way to generate a set of descriptive statistics for a dataset**

The mean, median, minimum and maximum are all "descriptive statistics." This means that they provide key information about the distribution of data. Here, we want to know the distribution of mercury in fish collected in 1998.

The function summary() estimates all these descriptive statsitics and more.
``` {r summary stats}
# The format of this summary() command is similar to other commands.
summary(fish_1998$mercury)
```
1st Qu. and 3rd Qu. refer to the 1st and 3rd "Quartiles", which represent a quarter of the data. They tell you that 25% of the data is between that value, the Median, and either the Min. (1st Qu.) or Max. (3rd Qu.).

**Class Question:**

* Looking at all of these values, are there unsafe levels of mercury in local fish? Are there dangerous levels of mercury in most fish, or only some?


**Recap: How could we communicate these values?**

You now know how to get basic statistics in R, and that has already given a lot 
of helpful information for tribal fishery managers. However, summary stats can't 
answer every question, and as a list of numbers they would be very difficult to 
communicate to the K'avi community. It would be much more helpful to visualize the data in a graph. 

# PART 3: Working with ggplot to generate a graph of mercury concentrations

**First we want to visualize the amount of mercury in *all* the fish**

We can do this my creating graphs!

**NOTE TO GEORGIA: Should we provide a commented out example for Georgia to walk through (ggplot(data, aes(x,y) etc.)) ?** 

If you run the code chunk below, it will make a graph of all the 1998 mercury data. We want to see the distributions of mercury in the fish to learn how common it is for fish to have mercury levels above the EPA's recommended safe level.

This code has a lot of details that we will work through later. For now, you can either just run the code chunk (by clicking the small green arrow in the upper right corner of the grey area), or you can explore the code line by line. We've provided some comments to help you out. [This cheatsheet](https://images.datacamp.com/image/upload/v1666806657/Marketing/Blog/ggplot2_cheat_sheet.pdf) will also give you some tips. 


```{r ggplot, Ribbon Plot, mercury}
#For this plot, we want to see the distributions of mercury in the fish and how many of them are above the EPA's recommended levels
#ggplot is the function that we are using to create a plot. aes() is used to tell ggplot what we want to see on each axis, in this case x= 'fish' and y= mercury. x="fish" indicates that the x axis doesn't have a variable and is titled fish. This allows us to look at the variation in lengths of fish independent of another variable.
ggplot (fish_1998, aes(x= 'fish', y= mercury))+
#Since ggplot is just a function to create plots, we need to further specify what kind of plot it should create. We can do this by using the geom_jitter function, which will create the Ribbon plot seen below. We can also add a few more parameters to the graph, such as changing the width of the points (width = 0.1).
#The jitter plot is a variation on the scatterplot, but it separates the data so that we can distinguish how many points are above the unsafe mercury level. To see the difference, try geom_point () and then geom_jitter ()!
  geom_jitter(width = 0.1)+
#We can also add shapes to the plot. Here we added a line to show where  the EPA's guideline for unsafe levels of mercury in fish would fall. To create  this we used the geom_hline function which simply creates a horizontal line at a y-intercept that you set. We can also adjust the color and appearance of the line with color = 'red'.  
  geom_hline(yintercept = 4, color= "red")+
#This allows us to relabel the x-axis
  xlab("fish")+
#This allows us to relabel the y-axis
  ylab("mercury (mu/g)")+
#This allows us to title the graph
  ggtitle("1998 Mercury Distribution")+
#This adjusts the non-data components of the plot, so the titles, labels, background, and legends among other aspects. 
  theme_minimal()

```
**Class question:**

* What does this distribution tell us about the mercury levels in the fish? 

Next, Look back at the min, max, median, and mean values that were calculated above. Where would these values fall on this graph? Are the mean and median values above or below the level suggested by the EPA?

We can overlay a "boxplot" on the previous ribbon plot. Boxplots are a way to show the mean, quartiles, minimum and maximum of a distribution. You can google "boxplot" to learn more.

Like last time, you can just run this code if you want. We want to focus on the results. However, we've provided comments to help you understand each line of the code, if you want.
```{r, Ribbon and Box Plot}
# These lines are almost the same as the last code chunk, 
# but we added a '+' at the end, to add some new pieces to the graph.
ggplot (fish_1998, aes(x= 'fish', y= mercury))+
geom_jitter(width = 0.1)+
# We will overlay a boxplot on the jitter plot. 
# We can adjust its color as we did above,
# We use alpha, to adjust the opacity of the boxplot.
# This lets us see the ribbon plot beneath it.
geom_boxplot(alpha = 0.4, color="dodgerblue2")+
# As before, we add a line to represent the EPA's recommended safe mercury ingestion level
geom_hline(yintercept = 4, linetype= "dashed", color= "red", size=1)+
# The rest of the code is the same as the last code chunk:
  xlab("fish")+
  ylab("mercury (mg/kg)")+
  ggtitle("1998 Mercury Distribution + Box Plot")+
theme_minimal()
```

You have now seen what the mercury distribution in local fish looks like and connected it to the summary statistics that were calculated above!

**Class Questions:**

* From looking at this graph, are there unsafe levels of mercury in local fish? 
  If so, is it in most fish, or only some? Did this graph tell you anything different than the summary       statistics, or did it just tell you the same information in a different way?

* In 1998, was it safe for members of the K'avi community to eat fish from these waters?
* How comfortable would you have been in 1998 about eating a fish that you caught?

Methylmercury is known to build up in fish as they age: The oldest fish tend to have the highest levels of mercury. Older fish also tend to be bigger, so maybe if we know the relationship between fish size and fish mercury levels, that could give us useful guidance for which fish were safe to eat back in 1998.  

We'll make another graph of the same dataset. This time we'll make a scatterplot of both the mercury level and the length of each fish. A scatterplot shows the data as points in two dimensions (length and mercury, here. 

Again, you can just run the code (with the green arrow), but we've also provided comments to help you learn how it works. Our goal here is to let you see the kings of graphs you can make in R and to help interpret the safety of the fish from 1998.
```{r ggplot, Day1 1998 Scatter Plot}
# This line defines a plot 'p1' as a ggplot(). 
# The ggplot() code is a little different from the ribbon plot.
# Now, we are plotting two columns of the dataset: 
# length will be on the (horizontal) x-axis
# mercury will be on the (vertical) y-axis
LengthMercury1998 <- ggplot(fish_1998, aes(x = length, y = mercury)) +
# geom_point() makes a scatterplot.
# We are going to plot the data points in blue 
# and a little bigger than the default "size = 2"
  geom_point(color = 'blue', size = 2) +
# Now we're going to write labels for the two axes, and give the graph a title
  xlab("length (cm)")+
  ylab("mercury (mg/kg)")+
  ggtitle("Mercury (mg/kg) vs. Length (cm) (1998)")+
# As before, we will add a line to represent the EPA's recommended mercury level
# This time, we'll make a red dashed line.
  geom_hline(yintercept = 4.5, linetype= "dashed", color = 'red')+
# We'll use the same theme as last time.
  theme_minimal()
#Because we assigned the graph a label, we have to call that label
LengthMercury1998
```
**Class Questions:**

* What does this tell us about the relationship between fish length and mercury levels? 
* Are there certain lengths of fish that have a higher amount of mercury?
* If you caught a fish in 1998, would this graph help you decide whether the fish was safe? 


--------------------------------------------------------------------------------

# DAY 2: 

We will now analyze data collected from the same monitoring sites in the K'avi 
community from 25 years later (2023). In light of data collected in 1998 and the
presentations tribal fishery managers made to communicate their findings to the 
community,efforts were made to clean up the heavy metals present in the water 
that were becoming concentrated in the local fish populations. 


Our goal today is to see whether the efforts made by tribal fishery managers were 
effective in reducing the amount of mercury present in the fish. We will do this
by comparing our findings from the 1998 dataset to our current findings in the 2023
dataset. 

# PART 1: Loading in and exploring the data

Like we did on Day 1 of the fish module, we first need to load in our datasets 
before we can explore them!

Before we begin, we check and set our working directory. First we will use the getwd()
command to figure out which working directory we are currently in. Then we will 
use "setwd()" to change our working directory to the directory where our data lives.
The easiest way to think of a directory is as a location on your computer where 
data and other files are located. Each of these directories can be accessed with 
their own unique file path, which is basically a list of different locations on 
your computer that becomes more specific as it continues! 

--- Hannah: Are we having both of these in the same document? If so then we do not need to set the working directory twice. If we are splitting them then I will change this to match the above ---

---Catherine: That's a good point, and based on what we discussed last class I 
think that both days' assignments will be in the same document so students can 
generate both plots easily and place them side by side?---

```{r set wd2}
# getwd()
# setwd()
# list.files()
```


## Load in packages 

We will be using the same packages as we used on Day 1. Remember to load packages, 
we used the command library(), with the name of the package in the parentheses. 

--- Hannah: See comment above ---

```{r load packages2}
library(tidyverse)

#We will need to use a new package today that we haven't installed before called 
#patchwork.
if(!require("patchwork")) install.packages("patchwork") 
library(patchwork)
```



## Now we will load in the dataset for the physical fish data

```{r fish2}
#fish <- read.csv("fish_2023.csv", header = TRUE, sep = ',')
load("fish_2023.RData")
fish_2023 <- fish_2023
```


## Now that we have loaded our data, we can begin to explore it to see what it 
contains. 

```{r View data pt. 1 - take 2}
View(fish_2023) 
```

## Recall from last class that there is another way to view only a small part 
of our data frame, using the argument "head()". 

*What would go in the parentheses of head() in order to get the output you want?

``` {r View data pt. 2 - take 2}
head() #Fill in the parentheses 
```

*What are the variables that this data set contains?

## It can also be helpful to explore the dimensions (size) of our dataset, so 
we know how big our sample size is. To do this, we will use the function "dim()".

*What would go in the parenthese of dim() in order to get the output you want? 

```{r explore data dimensions 2}
dim() #Fill in the parentheses with the name of the dataframe
```

##Recap: 

Now that we have successfully loaded in our data and explored its basic 
properties, we can now start to explore our data more deeply to understand how 
mercury levels in fish belly fat have changed from 1998 to 2023. 


# PART 2: Running some basic statistics! 

According to the EPA, the maximum safe level of mercury to be consumed is around
0.1 mg/kg/day. Tribal fishery managers are interested in how mercury levels in 
fish belly fat have changed since 1998 to 2023, now that the rivers have been 
cleaned. 

Below, we will analyze data points from the "fish_2" data set to try and answer 
these questions. We will start to do this by running some basic summary statistics 
on the fish_2 dataset. 


## What is the mean amount of mercury found in fish in 2023? Recall from last class 
that to find mean we use the command: mean(data_name$column_name). 

```{r mean amount of mercury 2}
mean(fish_2023$mercury_2023)

```


## What is the median amount of mercury found in fish in 2023? Recall from last class 
that to find median we use the command: median(data_name$column_name). 

*What would you fill in after the $ in the parentheses to get the output you want
(remember, we want to find the median amount of mercury in the fish).

```{r median amount of mercury 2}
median(fish_2023$)
```

*How do the mean and median values of mercury compare to the EPA's standard? 


## Now figure out what the maximum and minimum amount of mercury was found in fish
##in 2023. 
``` {r max and minimum amount of mercury 2}

```
 
## Recap: How could we communicate these values? 
By running these summary statistics, we can see that mercury levels in fish caught 
in 2023 are lower than they were in fish that were caught in 1998, which is 
good news! However, tribal fishery managers think that in order to truly understand
how mercury levels in fish have changed since 1998, it would be best to visualize
these data points in graphs! 


# PART 3: Making graphs to better understand how mercury levels in fish have changed
since 1998 to 2023. 


# Like we did with the fish data from 1998, we want to visualize the distributions 
# of the fish by the amount of mercury that they have. We will make the same graphs
# as we did with the fish data from 1998 with the fish data from 2023. 

# For this plot, we want to see how many fish are above the EPA's recommended 
levels for mercury consumption. 

##Below is code to make the plot! Don't worry about the code starting with "#",
this contains details on how to make the plot. 

```{r ggplot, Ribbon Plot, mercury - fish_2023}
#ggplot is the function that we are using to create a plot. aes() is used to tell ggplot what we want to see on each axis, in this case x= 'fish' and y= mercury. x="fish" indicates that the x axis doesn't have a variable and is titled fish. This allows us to look at the variation in lengths of fish independent of another variable.
ggplot (fish_2023, aes(x= 'fish', y= mercury))+
#Recall that we need to tell ggplot what kind of plot we want to make! We want to make a scatterplot, but 
#we want a little more space between our points than a normal scatterplot, so we will call for "geom_jitter()".
#Width helps us set up how far apart we want the point to be. 
  geom_jitter(width = 0.1)+
#To aid in our understanding of where fish with dangerous amounts of mercury fall, 
#we will add a horizontal line on our plot signifying this "unsafe" level. To do 
#this we will use the command geom_hline(). 
  geom_hline(yintercept = 4, color= "red")+
#This allows us to relabel the x-axis
  xlab("fish")+
#This allows us to relabel the y-axis
  ylab("mercury (mu/g)")+
#This allows us to title the graph
  ggtitle("2023 Mercury Distribution")+
#This adjusts the non-data components of the plot, so the titles, labels, background, and legends among other aspects. 
  theme_minimal()

```
#What does this distribution tell us about the mercury levels in the fish in 2023? 

#Look the min, max, median, and mean values that were calculated above. 
#Where would these values fall on this graph? Are the mean and median values 
#above or below the level suggested by the EPA?

#We will now make the same plot again, but layer another type of plot (a boxplot)
#on top of it. The boxplot will help us see where our summary statistics values
(mean, minimum, and maximum) directly fall on our data. 
```{r, Ribbon and Box Plot}
#Recall that to make a plot, we first must run the line of code for ggplot(data, aes(x = variable, y = variable)).
ggplot (data = fish_2023, aes(x= 'fish', y= mercury))+
#We specify that we want to create a ribbon/jitter plot and adjust its width
geom_jitter(width = 0.1)+
#We also want to overlay a boxplot on the jitter plot. We can adjust its color and 
#its transparency so that we can see our individual data points at the same time. 
geom_boxplot(alpha = 0.4, color="dodgerblue2")+
#Again, we add a line to represent the EPA's recommended safe mercury ingestion level
geom_hline(yintercept = 4, linetype= "dashed", color= "red", size=1)+
#Label the x-axis
  xlab("fish")+
#Label the y-axis
  ylab("mercury (ug/g)")+
#Title the graph
  ggtitle("2023 Mercury Distribution + Box Plot")+
#Adjust the theme of the graph
theme_minimal()
```

#You have now seen what the mercury disriubtion of the fish in the lake looks like 
and connected it to the summary statistics that were calculated above. But, this 
graph doesn't tell us anything about how mercury levels and the length of the fish 
in the lake are related. We know what the data in 2023 look like, but we still haven't
compared it to the data from 1998 in a graph to see if the
lake clean up was actually successful. 

#Tribal fishery managers think that we should plot the data from fish_1998, and plot the 
#data from fish_2023, and put these plots side by side to compare them. These next steps will 
#accomplish that! 

#Again, don't worry about the code below, just think about the information that we can take from it. 
#The code below will download our fish_1998 data, and make the same plot as we did last class. 
```{r load fish_1998, ggplot, Scatter Plot}
load("fish_1998.RData")
fish <- fish_1998


#We will call ggplot, tell it the data we want to use, and assign different variables 
#to the x and y axis. 
  p1<- ggplot(fish_1998, aes(x = length, y = mercury)) +
#We then need to tell ggplot what kind of graph we are making, so we add geom_point for a scatterplot. 
#We will set the color to blue and the size of the point to 2. 
  geom_point(color = 'blue', size = 2) +
#Label the x-axis
  xlab("length (cm)")+
#Label the y-axis
  ylab("mercury (ug/g)")+
#Title the graph
  ggtitle("Mercury (ug/g) vs. Length (cm) (1998)")+
#We add a dotted line to represent the EPA's recommended mercury level and then adjusted the color and type of line
  geom_hline(yintercept = 4.5, linetype= "dashed", color = 'red')+
#Adjusted the theme of the graph
  theme_minimal()
#Because we assigned the graph a label, we have to call that label. This label 
#makes our plot an object. Objects in R are things like plots that are given a 
#short, specific name that are easy to call for in our code when we want to do 
#things with it (like view it, or put 2 graphs next to one another). 
p1
```


We'll practice using ggplot again to visualize fish 
data from the fish_2023 dataset, and compare our results to those we generated 
with the fish_1998 dataset. 

## Recall from last class the steps to make a plot with ggplot:
```{r ggplot, Scatter Plot 2}
#Call "ggplot" followed by the "aes() argument. Which data would 
  #go on the x and y axis? 
p2<- ggplot(fish_2023, aes(x = length, y = mercury)) +
#Now it's time to specify which plot we want ggplot to make. A scatterplot
#would be good to visualize our data, so we will call "geom_point()". In the 
#parentheses, we can specify extra details about how we want the graph to appear.
#Here, we will set the color to green, and the size of the points to 2. 
  geom_point(color = 'green', size = 2) +
  #Label the x-axis
  xlab("length (cm)")+
#Label the y-axis
  ylab("mercury (ug/g)")+
  ggtitle("Mercury (ug/g) vs. Length (cm) (2023)")+
#To truly understand if efforts to clean up the waterways made a difference in 
#mercury concentrations in the fish, we will again place a dashed line on the 
#graph where the EPA deems the unsafe level of mercury in fish would be. We will
#do this by calling geom_hline, then setting the y-intercept of where we want the
#line to be placed, set the "linetype" and set the "color" of the line. 
geom_hline(yintercept = 4.5, linetype= "dashed", color = 'red')+
#We will use our same theme
  theme_minimal()

#To see our plot, we need to call it's object name, which is p2!
p2
```

## Tribal fishery managers believe that the best way to see if efforts to clean 
## the waterways were successful would be to put our plot from 1998 next to 
## our plot from 2023.

```{r putting plots side by side}
p1 + p2 #plot from 1998 next to plot from 2023 
```


*What is the safest large size fish to eat based on the 2023 dataset? How does this compare
to the safest large size fish that could be eaten based on the 1998 dataset? 

*Based on the distribution of data in our plot from the fish_2023
dataset compared to how our plot looked with the fish_1998 dataset, were the clean-up
efforts successful? If so, how can you tell? 







