---
title: "Mod2_water"
output: html_document
date: "2023-11-14"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DRAFT: Background for Module 2: Water Quality

Water sustains life on our planet and water is sacred to Native Americans. "Water quality" captures many components of the chemistry, biology, and physics of fresh water. In this module, we will study data about two of these components, water temperature (T) and dissolved oxygen (DO). 

Bull trout are a culturally important, but threatened species of fish native to the U.S. Northwest and the Canadian Rockies. They provide an example of how water quality affects Indigenous culture, because they require water temperatures below 14 degrees C and dissolved oxygen levels above 6.2 mg/L. Such cold clean water conditions are growing scarcer due to worsening water quality caused by climate warming and pollution.

[#JMcnote: We need to check these values.]

In this module, we will use data about water quality trends to help the K'avi Fish and Wildlife department decide which of two streams would be best for a reintroduction of bull trout on tribal land. The tribe only has funding to bring the bull trout back to one stream and they have narrowed the choices of possible locations for the reintroduction to two streams, where temperature and D.O.have been monitored for about 15 years. Neither site is perfect, but they are both pretty good options. Your job is to analyze the temperature and D.O. data to help the tribe decide which is the best site. 

Stream A is 

analyze water temperature and dissolved oxygen data the K'avi collected from two streams. The question the K'avi fish and wildlife biologists have to answer

VIGNETTE OUTLINE:
Set up: 
- Bull trout: culturally important and threatened species.
- Thresholds of low D.O. and high T that are dangerous for bull trout (from lit)
- Goal: Picking between 2 streams for trout introduction.
-- Stream A is fed by a nearby groundwater spring.
-- Stream B is outflow from a mountain lake. 


-------
=======
VIGNETTE:
**Some questions/thoughts [from CRA]:
- I know in class we discussed the angle for the vignette is that tribal land 
managers are interested in how water quality is changing due to climate change.
This is also relevant based on Georgia's EJDS 2023 example 
- Is there a way to frame this that would make it more relevant to students? 
- Maybe introduce a reason - tribal communities use the water in X way 
(fishing etc.), this is why they're concerned
  - Could lead into why temperature and DO are important (fish kills?)
  - DO levels being low could be an important factor, and raising temperatures
  causes less oxygen to be dissolved in water. 
  
>>>>>>> 6a5ee821ae1c316b9e69ecc940debbde9cc596ff
Notes from JMc to Catherine about data:  
Overall Data qualities: 
(1) We should have average *annual* T and DO data (because Temp and DO will cycle seasonally, which will both make these data harder to simulate and also be more difficult to visualize trends over time)
(2) Stream A should have a shorter observational length (like 2012 - 2022), than Stream B (say 2007 - 2022). This is because we want some NA values that we will learn to delete later.]
(3) The data should be in wide form (described in summary below), which will allow us to "tidy" the data by pivoting to long form data.

Stream A Data qualities: 
  D.O. is on the low side (occasionally sneaking into the danger zone), but Temp is OK. No trend in either, because the groundwater spring keeps Temp stable, and DO is a function of Temp.

Stream B Data qualities: 
  On average, D.O. and Temp are better than Stream A, but the trend is towards higher T and lower D.O. (because climate change is warming the lake water and D.O. declines as T increases)

Summary of fake data set:
  - 5 Columns: headings are "Year", "StreamA_Temperature". "StreamA_DO", "StreamB_Temperature". "StreamB_DO"
- 15 Rows: 2007 to 2022
- Values are as described above, with NA for Stream A from 2007 to 2011
-------
  
Overview of upcoming steps:
1. Import data, pivot to tidier long format & removing NAs, Calculate summary stats for both sites. 
Q: Which site looks better? A: Site B, because it has higher D.O. and lower T.

2. Make a scatter plot of Temp & D.O. from both sites. Plot sites with different colors. 
Add lines (or overlay a box) identifying the T and D.O. thresholds. 
Q: Which site looks better? A: Site B, again, because more points are outside of the "danger zone."

3. Now let's make a new kind of graph: time series of changing T and D.O. from the two sites.
First graph is Temperature from both sites. Second graph is DO from both sites 
(use the same colors for each site as in step 2). 
NOTE: THIS IS REVISED FROM 11/20: AVOID THE DOUBLE AXIS COMPLICATIONS AND EASIER 
TO COMPARE TRENDS ACROSS SITES.
Place graphs side-by-side. Q: Which site looks better now? A: Maybe Site A, because 
even though it has more days when the water quality dips into the danger zone, 
it's not getting worse. Site B is better for now, but climate change seems to be 
moving it into  more unhealthy conditions.


In this module, there are four main coding goals: 
 1. We will explore how water quality data can help communities make decisions 
to protect culturally important species.

2. We will learn how to rearrange data into a "tidy" format that makes graphing easier
and we will learn how to "debug" code, in other words, we'll learn how to fix common 
problems that we see when using real world data. 

3. We will learn how to separate data into meaningful categories in R based on 
similarities. This is known as subsetting. 

4. We will learn how to make graphs of data over time. These are called 
"time series" graphs.  
  


--------------------------------------------------------------------------------
#INSTRUCTOR'S VERSION: Learning goals 

1. Students will review the steps to set a working directory and load in data from 
an extrenal data file. 

2. Students will review basic commands in R to explore their data before they 
begin to analyze it. 

3. Students will review how to load in packages. 

4. Students will review basic summary statistic commands in R to pull the mean, 
median, max, and min values for different variables in the dataset. 

5. NEW SKILL: students will learn how to handle common code bugs, such as NAs. 

6. NEW SKILL: students will learn how to subset data in R based on scientifically 
relevant groups. 

7. NEW SKILL/review: Students will review how to make a graph using ggplot. 
Students will learn how to make a new kind of plot in ggplot (line graph),
in addition to the scatterplot and ribbon plot they learned about in Module 1 fish.
This will be framed in understanding how we should present data in a meaningufl 
way to any community, 

--------------------------------------------------------------------------------
#PART 1: Loading in and exploring the data 

Like we did in Module 1, before we being to analyze our data, we need to do a few
steps first.

#Clearing Environment 

Before we start to do anything in R, we should clear our environment. As a reminder,
your environment is where R stores all the objects and data you were working with
since the last time you worked in R. See the code chunk below on how to clear 
your environment. 

```{r}
rm(list = ls())
```


#Setting the Working Directory 

We will now set our working directory. As you may remember, setting our working 
directory tells R where to find the files we want to work with on our computer. 

[JMc comment: We should decide as a class whether we want to set working directories this way, 
or just put the data files in the same directory as this .Rmd file. 
When you knit an R Markdown it looks for the data in that directory.
See this link, too: https://github.com/yihui/knitr/issues/277#issuecomment-6528846
]


```{r}
current_path = rstudioapi::getActiveDocumentContext()$path
setwd(dirname(current_path))
```


#Load in packages 

Next, we will load in packages, to get the extra tools we need for this module 
loaded in R! We will load in the tidyverse package. 

```{r}
library(tidyverse)
```


#Load in our data! 

Now it's time to load in our data! In this scenario, temperature data 
and dissolved oxygen data were collected and stored in two separate files. 
Therefore, we will need to load in both files in order to use them. 

(NOTE: this may change if data is not in a 
csv file)

```{r}
#We will first load in our data for water temperature. 
stream_temp <- read.csv("streams_temperature.csv", header = TRUE, sep =",")
```

```{r}
#Now we will load in our data for dissolved oxygen concentrations.
stream_DO <- read.csv("streams_DissolvedOxygen.csv", header = TRUE, sep =",")
```

#Viewing the data 

Now that our data is loaded into R, we can explore the data using the view and 
head command. 

```{r}
View(stream_temp)
```

```{r}
head(stream_DO)
```

*What do you notice is different between using the command "View" and the command
"head"? Which one do you think is more useful? 

It may also be helpful for us to understand how big our dataset is. We can 
get this number by using the command "dim".

```{r}
dim(stream_temp)
dim(stream_DO)
```

*How many data entries are in this dataset for water temperature? 

*How many data entries are in this dataset for dissolved oxygen? 


RECAP:
You now reviewed ways to set up your R environment, read in data, and explore 
your data! These steps are all important for when we analyze our data later in 
this module. 


--------------------------------------------------------------------------------
#PART 2: Pivoting data, debugging data, and running some basic summary statistics

[CRA: from here up until line 302 are edits I made on 11/28 per the new data]

Tribal land managers are interested in how the temperature and dissolved 
oxygen of the water on their lands have changed over the last several years. 
They are also interested in learning about the mean, median, minimum, and maximum
values of temperature and dissolved oxygen in water they have in their data. 
These values will help them understand how current values compare to the rest of
the values in their data. 


As of now, our data for water temperature for Stream A and B exists in a separate
data frame from the dissolved oxygen concentrations for Stream A and B. For
tribal land managers to truly understand what's going on, they need to analyze
water temperature and dissolved oxygen together. To do this, we need to change 
how our data is set up in each data frame and then merge these data frames together. 


To modify our data we need to do something called "pivoting". In our current 
data for water temperature, we have columns for the year the sample was collected,
the temperature for the samples from Stream A, and the temperature of the samples in 
Stream B. Instead, we want columns for the year, the site where the sample was collected, 
and a column for all the temperatures. 

To do this we will use a command from tidyverse called "pivot_longer". See the 
code chunk below for more explanation! 
```{r}
# this first line of code tells R that we want to
# re-save the streams_temp dataset with the changes we make to it with pivot_longer. 
stream_temp <- stream_temp %>%  
  #we use the pivot_longer code to tell R which columns we want to change.
  pivot_longer( 
    #cols = tells R which columns we want to modify
    cols = c(StreamA, StreamB), 
    #names_to tells R that we want the names of the columns we 
    #specified in the line above to be entries in a new column we're creating 
    #called "site". 
    names_to = "site", 
    #values_to = tells R that we want the values from the columns in col = 
    #to be moved into a new column we're making called "temperature_C". 
    values_to = "temperature_C" 
  )
```

To check our work, we can view streams_temp: 
```{r}
View(stream_temp)
```


Great! Our code worked. Now let's do the same for the dissolved oxygen data set.
We want to change it in the same way we changed the temperature data set above. 

```{r}

stream_DO <- stream_DO %>%
  pivot_longer(
    cols = c(StreamA, StreamB),
    names_to = "site",
    values_to = "dissolved_oxygen"
  )
```

Again, we should check our work to make sure that our data looks the way we want
it to. We want there to be three columns, one for the year the sample was collected,
one for the site name, and one for the dissovled oxygen values (mg/L). 

```{r}
View(stream_DO)
```

Good work! This data frame looks the way we want it to now as well. 

Next, we need to merge these two data frames together into one, so we can analyze 
trends in temperature and dissolved oxygen at the same time for both streams. 


To do this, we will use a command in tidyverse called "left_join". 
Left join means we want to put one data frame next to the the other. In other words, 
we want to match up the rows in the data frame, so it becomes wider, not longer. This is 
because we want to preserve the columns from the two data frames we are combining. 
```{r}
streams <- left_join(stream_temp,stream_DO)
#Here, we are using left_join to put the streams_temp data frame next to the streams_DO
#data frame.
```


Great! Now let's check to make sure that we successfully merged our two data frames.
```{r}
View(streams)
```


Great job data wrangling! We can now continue to analyze our data. 


#

[#JMc note to CRA @ 11/28 8pm: I think code below here needs updating.]
```{r}
mean(water$temperature)
```

Ah! Our code didn't work. R gave us an error message that says: "". This means 
we have some missing chunks of data in our data set that R doesn't know how to 
work with. 

Sometimes when collecting data, it's possible that we don't have all complete 
data sequences for all samples in our data set. While this is very normal, as we can
see, it can make running commands in R a little bit tricky.

These missing data are entered as NA (not available) in R automatically.
In this section, we'll go over some easy ways to fix this problem in our "water"
data set by telling R to ignore the NAs. 

```{r}
#JMc We're going to change this to dplyr 
# This will help 
# https://dplyr.tidyverse.org/
# JMc e.g., https://www.youtube.com/watch?v=8SGif63VW6E
na.omit(water)
```

Great! Now let's try to run our command to get the mean for temperature again. 


```{r}
mean(water$temperature)
```

Our code worked! Now you've learned how to handle NAs in a dataset in R. Let's
carry on with our summary statistics.

*Your turn! How would you figure out what the mean amount of dissolved oxygen 
is in the water across all samples? 

```{r}
#Write your code here!

```


#What is the median amount of temperature across all samples? 

```{r}
median(water$temperature)
```


#What is the median amount of dissolved oxygen across all samples? 

```{r}
#Write your code here! 

```

#What is the maximum and minimum temperature in our data?

```{r}
max(water$temperature)
min(water$temperature)
```

#Your turn: figure out the minimum and maximum amount of dissolved oxygen recorded
in our dataset? 

```{r}

```
