---
title: "Mod2_water"
output: html_document
date: "2023-11-14"
---

VIGNETTE:
**Some questions/thoughts:
- I know in class we discussed the angle for the vignette is that tribal land 
managers are interested in how water quality is changing due to climate change.
This is also relevant based on Georgia's EJDS 2023 example 
- Is there a way to frame this that would make it more relevant to students? 
- Maybe introduce a reason - tribal communities use the water in X way (fishing etc.), 
this is why they're concerned
  - Could lead into why temperature and DO are important (fish kills?)
  - DO levels being low could be an important factor, and raising temperatures
  causes less oxygen to be dissolved in water. 



In this module, there are three main coding goals: 
1. We will explore data related to water quality on tribal lands 
2. We will learn how to "debug" code, in other words, we'll learn how to fix 
common problems that we see when using real world data that can sometimes make 
coding tricky if not fixed. 
3. We will learn how to separate data into meaningful categories in R based on 
similarities. This is known as subsetting. 


--------------------------------------------------------------------------------
#INSTRUCTOR'S VERSION: Learning goals 

1. Students will review the steps to set a working directory and load in data from 
an extrenal data file. 

2. Students will review basic commands in R to explore their data before they 
begin to analyze it. 

3. Students will review how to load in packages. 

4. Students will review basic summary statistic commands in R to pull the mean, 
median, max, and min values for different variables in the dataset. 

5. NEW SKILL: students will learn how to handle common code bugs, such as NAs. 

6. NEW SKILL: students will learn how to subset data in R based on scientifically 
relevant groups. 

7. NEW SKILL/review: Students will review how to make a graph using ggplot. 
Students will learn how to make a new kind of plot in ggplot (line graph),
in addition to the scatterplot and ribbon plot they learned about in Module 1 fish.
This will be framed in understanding how we should present data in a meaningufl 
way to any community, 

--------------------------------------------------------------------------------
#PART 1: Loading in and exploring the data 

Like we did in Module 1, before we being to analyze our data, we need to do a few
steps first.

#Clearing Environment 

Before we start to do anything in R, we should clear our environment. As a reminder,
your environment is where R stores all the objects and data you were working with
since the last time you worked in R. See the code chunk below on how to clear 
your environment. 

```{r}
rm(list = ls())
```


#Setting the Working Directory 

We will now set our working directory. As you may remember, setting our working 
directory tells R where to find the files we want to work with on our computer. 

```{r}
current_path = rstudioapi::getActiveDocumentContext()$path
setwd(dirname(current_path))
```


#Load in packages 

Next, we will load in packages, to get the extra tools we need for this module 
loaded in R! We will load in the tidyverse package. 

```{r}
library(tidyverse)
```


#Load in our data! 

Now it's time to load in our data! (NOTE: this may change if data is not in a 
csv file)

```{r}
water <- read.csv("water_quality.csv", header = TRUE, sep =",")
```

#Viewing the data 

Now that our data is loaded into R, we can explore the data using the view and 
head command. 

```{r}
View(water)
```

```{r}
head(water)
```

*What kind of variables are in our dataset for water quality on the K'avi lands? 

*Are there some variables in particular that you think would be important for 
tribal land managers to focus on to understand how climate change is affecting 
water quality on their lands? 

It may also be helpful for us to understand how big our dataset is. We can 
get this number by using the command "dim".

```{r}
dim(water)
```

*How many data entries are in this dataset? 


## Recap 
You now reviewed ways to set up your R environment, read in data, and explore 
your data! These steps are all important for when we analyze our data later in 
this module. 


--------------------------------------------------------------------------------
#PART 2: Debugging data, and running some basic summary statistics

Tribal land managers are interested in how the temperature and dissolved 
oxygen of the water on their lands have changed over the last several years. 
They are also interested in learning about the mean, median, minimum, and maximum
values of temperature and dissolved oxygen in water they have in their data. 
These values will help them understand how current values compare to the rest of
the values in their data. 

#

```{r}
mean(water$temperature)
```

Ah! Our code didn't work. R gave us an error message that says: "". This means 
we have some missing chunks of data in our dataset that R doesn't know how to 
work with. 

Sometimes when collecting data, it's possible that we don't have all complete 
data sequences for all samples in our data set. While this is very normal, as we can
see, it can make running commands in R a little bit tricky.

These missing data are entered as NA (not availabe) in R automatically.
In this section, we'll go over some easy ways to fix this problem in our "water"
data set by telling R to ignore the NAs. 

```{r}
#to do this we can use na.omit 
na.omit(water)
```

Great! Now let's try to run our command to get the mean for temperature again. 


```{r}
mean(water$temperature)
```

Our code worked! Now you've learned how to handle NAs in a dataset in R. Let's
carry on with our summary statistics.

*Your turn! How would you figure out what the mean amount of dissolved oxygen 
is in the water across all samples? 

```{r}
#Write your code here!

```


#What is the median amount of temperature across all samples? 

```{r}
median(water$temperature)
```


#What is the median amount of dissolved oxygen across all samples? 

```{r}
#Write your code here! 

```

#What is the maximum and minimum temperature in our data?

```{r}
max(water$temperature)
min(water$temperature)
```

#Your turn: figure out the minimum and maximum amount of dissolved oxygen recorded
in our dataset? 

```{r}

```


**QUESTION: Do we want to introduce subsetting in the summary statistics section?
Or do we just want to keep it in the visualizations portion? If so, I would 
have some ideas on how to restructure the rest of this section. 





--------------------------------------------------------------------------------
#PART 3: Visualizing our data! 


