# Vignette

Members of the K'avi tribe have managed a forest on their lands for decades. Recently, tribal forestry managers have started collecting data. They want to analyze it to determine the growth rates and other statistics of the trees in the forest for a harvest. 

*Alternate vignette based on discussions in class*
The K'avi tribe would like to set up a location for managing and harvesting Douglas firs for lumber. A couple of different sites have been proposed for this new forestry plot and it is your job to determine which of the proposed sites would be best for douglas fir harvest. 

*We should flesh this out a little more.*

--------------------------------------------------------------------------------
# WORKSHOP: Learning goals (More for us as a team, rather than for SKC students)

The goal of the first day of the this forestry module is to have the students replicate what they did in the first 3 modules all at once and to do so more independently than previously. 

Things this module should include are: 

1. Min, max, median, mean review.  

2. Maximum sustained yield. 

3. More time-series exploration

4. More complex data visualization. 

--------------------------------------------------------------------------------
#PART 1-
**For the student version we will clear these segments so they have to come up with the code themselves but I have included the code for the instructor version. **
##Clearing the environment
```{r}
rm(list = ls())
```

##Setting the Working Directory
```{r}
# This isn't particularly necessary here because we aren't technically loading in external data but still probaly best practice.
current_path = rstudioapi::getActiveDocumentContext()$path
setwd(dirname(current_path))
```

##Load in Packages, what packages will you need to analyze and/or visualize this data?
```{r}
library(tidyverse)
```

## read in data

*Add in descriptions of the data in the context of the problem*
```{r}
site_1 <- read.csv("site_1.csv")
site_2 <- read.csv("site_2.csv")
site_3 <- read.csv("site_3.csv")
site_4 <- read.csv("site_4.csv")
```


## Exploring the data

What variables does the data you just loaded in contain? What do these variables 
mean in the context of forestry management? How could this data be used to 
determine which site is the best choice for forestry? 
Feel free to insert a code chunk below here to explore what the variables 
are in the data. 
*Have they had to insert a code chunk before? This seems like an important skill*

Now lets start to explore the differences between the sites with summary 
statistics. What statistics would help you explore the difference in 
productivity between the different sites? 

```{r}
# Use this chunk to do any summary statistics you feel would be helpful. 
```

Based on these summary statistics, which site seems like the best site so far? 
What other factors would you like to consider about each of these different 
sites that was not provided in this dataset? 

Create a tabble/tibble with the median age of the trees at each site
```{r}
summarize(site_1, site_1_median = median(site_1$age), site_2_median = median(site_2$age), site_3_median = median(site_3$age), site_4_median = median(site_4$age) )
```


## Combining the data

## Graphing the data
