# Vignette

Members of the K'avi tribe have managed a forest on their lands for decades.
Recently, tribal forestry managers have started collecting data. They want to
analyze it to determine the growth rates and other statistics of the trees in the
forest for a harvest. 

*We should flesh this out a little more.*

--------------------------------------------------------------------------------
# WORKSHOP: Learning goals (More for us as a team, rather than for SKC students)

The goal of the first day of the this forestry module is to have the students replicate what they did in the first 3 modules all at once and to do so more independently than previously. 

Things this module should include are: 

1. Min, max, median, mean review.  

2. Maximum sustained yield. 

3. More time-series exploration

4. More complex data visualization. 

--------------------------------------------------------------------------------
#PART 1-
**For the student version we will clear these segments so they have to come up with the code themselves but I have included the code for the instructor version. **
##Clearing the environment
```{r}
rm(list = ls())
```


##Setting the Working Directory
```{r}
# This isn't particularly necessary here because we aren't technically loading in external data but still nice to have.
current_path = rstudioapi::getActiveDocumentContext()$path
setwd(dirname(current_path))
```

##Load in Packages, what packages will you need to analyze and/or visualize 
this data?
```{r}
library(tidyverse)
```

# Data - 
```{r}
# This might require more explanation because I think this is the first time they're adding data from a package. 
trees <- Loblolly
```

Are there any aspects of the data that we need to clean up? Remove NAs?


Take a look at the data collected by the K'avi tribe, and don't forget to assign
the data to a variable to make it easier to work with.
```{r}
# This lets you see the beginning of the table
head(trees)
# This shows that there are no nas in the dataset. Do we want to add some? I am not sure this code has been presented yet. 
sum(is.na(trees))
```
What are the dimensions of the data? What kind of information does it provide us?

In the last module (Bison), we talked about creating a tibble (or R's version
of a table). Create a tibble for the statistics that you just calculated above.
```{r, tibble}
# Finding the dimensions
dim(trees)
# Making a tibble
tibble(dim(trees))
```


##Maximum Sustained Yield
**We're gonna need to do more research on this one. **
```{r}

```

Adding a new column to the data

##Graph
```{r, histogram}

```

#Adding the MSY to the graph
```{r}

```


What are the limitations of the MSY? What other questions could we answer with 
this data? 

##Maps?
```{r}

```



