# Vignette

Members of the K'avi tribe have managed a forest on their lands for decades.
Recently, tribal forestry managers have started collecting data. They want to
analyze it to determine the growth rates and other statistics of the trees in the
forest for a harvest. 

*We should flesh this out a little more.*

--------------------------------------------------------------------------------
# WORKSHOP: Learning goals (More for us as a team, rather than for SKC students)

The goal of the first day of the this forestry module is to have the students replicate what they did in the first 3 modules all at once and to do so more independently than previously. 

Things this module should include are: 

1. Min, max, median, mean review.  

2. Maximum sustained yield. 

3. More time-series exploration

4. More complex data vizualization. 

--------------------------------------------------------------------------------
#PART 1-

##Clearing the environment
```{r}

```


##Setting the Working Directory
```{r}

```

##Load in Packages, what packages will you need to analyze and/or visualize 
this data?
```{r}

```

# Data - 
```{r}
library(tidyverse)
trees <- Loblolly
```

Are there any aspects of the data that we need to clean up? Remove NAs?


Take a look at the data collected by the K'avi tribe, and don't forget to assign
the data to a variable to make it easier to work with.
```{r}

```
What are the dimensions of the data? What kind of information does it provide us?

In the last module (Bison), we talked about creating a tibble (or R's version
of a table). Create a tibble for the statistics that you just calculated above.
```{r, tibble}

```


##Maximum Sustained Yield
```{r}

```

Adding a new column to the data

##Graph
```{r, histogram}

```

#Adding the MSY to the graph
```{r}

```


What are the limitations of the MSY? What other questions could we answer with 
this data? 

##Maps?
```{r}

```



