---
title: "Module 4: Forest Management | DAY 1"
output: 
  html_document: default
date: "updated 1-6-2025"
---

# Introduction

Members of the K'avi tribe have managed a forest on their lands for decades. Recently, tribal forestry managers have started collecting data from a few different sites in the forest. They would like to set up a location for managing and harvesting Douglas firs for lumber. It is your job to analyze the data and determine which of the proposed sites would be the best for a harvest.

The K'avi want to choose a site that is the most productive, or in other words, has the greatest average board feet value for a given time period. Board feet are a unit of measurement that expresses the volume of logs and standing timber.

They have also found throughout the time that they have been caring for the forest, that there are many factors that affect the quality of a given site. Components such as altitude, aspect, soil, and rainfall, as well as the combination of these factors have an effect on the productivity of a site. 

1. Altitude affects productivity by shortening the length of the growing season and lowering the mean temperature. At high altitude, growth begins late and stops early, therefore making them less productive.

2. Aspect also affects productivity, as sites that face north, northeast, and east are more fertile (likely due to the sun's rays having less of an effect on the soil's dampness). Aspects that faced south to west had variable affects. Also, sites with south, southwest, and south aspects were less productive.

3. Soil must have the necessary substances to meet the conditions needed to grow. The most productive sites have deep, well-drained sandy loam soils, with the second-most being clay soils, and the least productive being gravel soils.

4. Rainfall, especially when it is abundant, greatly increases the tree growth at a site. The most productive sites have an annual precipitation of over 60 inches with less productive sites having an annual precipitation of less than 30 inches. 

Below are short descriptions of the sites the K'avi managers have been collecting data on:
  
1. Site 1: At an altitude of 2,000 feet, an aspect that faces south to west, and annual precipitation of less than 40 inches
2. Site 2: Faces northeast, has well-drained, sandy loam soil and an altitude of 1,500 feet.
3. Site 3: Rainfall of less than 20 inches annually, a southern aspect, and gravel soil
4. Site 4: Sandy loam soil, an altitude of 1,600 feet and 65 inches of rainfall annually.

**Overview of Module**

* Exploring the data 

* Summary statistics

* Choosing a site

* Combining the data

* Calculating statistics from the combined data

* Creating a table 

* Graphing

* Recap

**Learning goals for this module:**

The goal of the first day of this module is to review what you have learned from the previous 3 modules and use that knowledge to answer questions about data more independently. More specifically:

1. Min, max, median, mean review

2. Exploring data sets

3. Combining data sets

4. Creating graphs

5. Using quantitative and qualitative data to make decision about which site is the most productive

6. Data visualization

## PART 1: Setting up and introducing the data

### Setting up the session

**Class Question:**

* What does the code below do? Add comments explaining its functions below:

```{r}
#
rm(list = ls())

# 
current_path = rstudioapi::getActiveDocumentContext()$path
setwd(dirname(current_path))

# 
library(tidyverse)
```

### Read in data

Below is the data that the forestry managers have been collecting from the different site:
```{r, load in the data}
site_1 <- read.csv("site_1.csv")
site_2 <- read.csv("site_2.csv")
site_3 <- read.csv("site_3.csv")
site_4 <- read.csv("site_4.csv")
```

## PART 2: Exploring the Data

Now that you've loaded in the data from the different sites,insert a code chunk below to explore what the variables are in the data:

```{r, look at the data}
#Use this chunk to look at the data above

head(site_1)
head(site_2)
head(site_3)
head(site_4)

#Could also use dim, glimpse, view....
```

**Class Questions:**

* What are the dimensions of the data from each site? 
* What variables does the data you just loaded in contain? 
* What do these variables mean in the context of forestry management?
* How could this data be used to determine which site is the best choice for forestry? 

### Summary statistics
Now lets start to explore the differences between the sites with summary statistics. 

**Class Question:**

* What statistics would help you explore the difference in productivity between the different sites? 

You can use the following chunk to perform whichever summary statistics you would find helpful. 

```{r, summary statistics}
# Use this chunk to do any summary statistics you feel would be helpful. 

```

**Class Questions:**

* Based on these summary statistics, which site seems like the best site so far? 
* Is there any other information you would like to consider about these sites that was not provided in this data set?

## PART 3: Combining the sites

Now that you've loaded in and explored that data from each of the sites, it will be useful to combine it in one large data set. We will use `mutate()`, similar to the previous module, to create a new column to identify each site. 

Then we will combine the rows using `rbind()`. This function combines all the data vertically, but in order to do so, all columns must have the same name. 

```{r, combining the data}
# create new column site_ID 
site_1_new <- site_1 %>%
  mutate(site_ID = 1)
site_2_new <- site_2 %>%
  mutate(site_ID = 2)
site_3_new <- site_3 %>%
  mutate(site_ID = 3)
site_4_new <- site_4 %>%
  mutate(site_ID = 4)

# combine rows with rbind
forest<- rbind(site_1_new, site_2_new, site_3_new, site_4_new)

# preview it 
head(forest)
```
### Calculate site averages 

Now calculate the mean board feet at each site from the new combined data. Similar to previous modules, use `group_by()` and `summarize()`

```{r, calculating board_feet}
#Use this chunk to calculate the mean board_feet from the combined data
forest %>% 
  group_by(site_ID) %>% 
  summarize(mean_bf = mean(board_feet))

```

Again using `group_by()` and `summarize()`, create a table/tibble with the median age of the trees at each site
```{r, table}
#Use this chunk to create the table

  
```

## PART 4: Graphing the data

In order to compare the different sites, we want to use the variable `site_ID` in our plots. Even though we know that it is a discrete value we are using as an ID, R recognizes this variable as a double which will cause issues when trying to plot it as a category. In order to fix this, we are going to save `site_ID` as a factor. Factors in R are categorical variables that can be stored in levels. 

```{r factor}

forest$site_ID <- as.factor(forest$site_ID)

```

### Boxplot

Create a boxplot with the `forest` dataframe showing `site_ID` on the x-axis, and `board_feet` on the y-axis.

```{r boxplot}

```

### Scatter plot 

Create a scatter plot with the `forest` dataframe of `board_feet` v. `age`, with `site_ID` as color or shape,  using `geom_point()` 

```{r, graph}


```

*Recap:*

You have now explored the data in many different ways. Based on the information from the summary statistics, graph, and descriptions of the sites, which site would you recommend as the best for a lumber harvest and why?
